---
title: "green_space_cleaning"
author: "Chris J. Martin"
date: "05/06/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)

library(janitor)    # for data cleaning
library(readxl)     # for readig excel files
library(visdat)     # for a quick look at data quality
```

## Processing the data

### Gardens data

Dataset downloaded from [ONS](https://www.ons.gov.uk/economy/environmentalaccounts/datasets/accesstogardensandpublicgreenspaceingreatbritain) on 5th June 2021. Both gardens space and public parks datasets are the April 2020 versions.

```{r}
# a convenience function for renaming columns
# I am sure there is a better way to do this!
replace_in_column_name <- function(df, pattern, replacement){
  df %>% 
    rename_with(.fn = ~ str_replace(.x, pattern, replacement))
}

# read in and clean data
gardens <- read_xlsx(
  "data/osprivateoutdoorspacereferencetables_edited_for_import.xlsx",
  sheet = "MSOA gardens", 
  skip = 1) %>% 
  
  # make variables names more conistent
  janitor::clean_names() %>% 
  
  # focus on variables of interest 
  select(country_code:msoa_name,
         ends_with("count") |
         starts_with("total")) %>% 
  
  # simplify variables names (now data by housing type has been removed)
  replace_in_column_name("total_", "") %>% 
  replace_in_column_name("private_outdoor_space", "gar") %>% 
  replace_in_column_name("address", "ad") %>% 
  replace_in_column_name("adress", "ad") %>%
  replace_in_column_name("adresses", "ads") %>% 
  replace_in_column_name("percentage", "perc") %>% 
  replace_in_column_name("_m2", "") %>% 
  replace_in_column_name("average", "ave")

# display data for quick visual checks
gardens
  
  
```

```{r}
# get a quick visual summary of data types and the amount of missing data
gardens %>% 
  visdat::vis_dat()

gardens %>% 
  visdat::vis_miss()
```

### Population data

ONS provide [MSOA level population data](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/middlesuperoutputareamidyearpopulationestimates). Here I use the most recent release (mid 2019).

```{r}
msoa_pops <- read_xlsx(
  
  "data/SAPE22DT4-mid-2019-msoa-syoa-estimates-unformatted.xlsx",
  sheet = "Mid-2019 Persons",
  skip = 4) %>% 
  
  # process variable names for conistency
  janitor::clean_names() %>% 
  
  # select minimal number of columns
  select(msoa_code, population = all_ages)

msoa_pops %>% 
  visdat::vis_miss()
```

### Indicies of multiple deprivation

ONS released the 2019 IMD data at LSOA scale. MySociety have produced [IMD at various other scales](https://research.mysociety.org/sites/imd2019/about/) including the MSOA scale.

```{r}
imd_2019 <- read_csv("data/imd2019_msoa_level_data.csv")
imd_2019
```

### Transforming the data for plotting

Overall average = 2.4 (2020 see [here](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/families/bulletins/familiesandhouseholds/2020))

2017-18 was the most year I could find on [average occupancy figures broken down by house/flat from MCHLG](https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/817286/EHS_2017-18_Households_Report.pdf):

-   House = 2.5 people

-   High-rise flat = 1.9

-   Mid-rise flat = 1.8

More details [here](https://www.gov.uk/government/statistics/english-housing-survey-2017-to-2018-households).

So for now, I'll make a conservative assumption that the average occupancy for a flat is 1.8 (i.e. the lower of the mid and high rise figures above).

**I looked at trying to estimate more accurately, but the figures for the number of dwellings per block are not available.**

2020 estimates of the numbers of high and mid rise flat can be found [here](https://www.gov.uk/government/publications/building-safety-programme-estimates-of-ews1-requirements-on-residential-buildings-in-england/building-safety-programme-estimates-of-ews1-requirements-on-residential-buildings-in-england). These can be used to calculate an average flat occupancy rate.

-   12,500 blocks of high rise flats

-   77,500 blocks of mid rise flats

```{r}
# calculate occupancy rates
ave_house_occ <- 2.5
ave_flat_occ <- 1.8

gardens %>% 
  left_join(msoa_pops) %>% 
  mutate(house_ad_without_gar_count = houses_ad_count - houses_ad_with_gar_count,
         flats_ad_without_gar_count = flats_ad_count - flats_ad_with_gar_count,
         pop_without_gar = (ave_house_occ * house_ad_without_gar_count) +
           (ave_flat_occ * flats_ad_without_gar_count))
  
```

## An initial plot
